
\section{ADARA Network Protocol}
\label{network_protocol}

\subsection{Overview}

The objective of this protocol is to provide a simple, easy to parse
structure for the transmission and storage of the data associated with
an experimental run at the SNS. It must accommodate high data rate of the
neutron event information from the detector electronics while offering
the flexibility to describe and report data from a diverse and dynamic
collection of sample environment sensors and control systems.

While the format is primarily used for network transmission over TCP
streams, certain packets will also be used over UDP connections when
communicating with embedded systems. Other packets are defined to enable
partial recovery from disk errors at inopportune locations in the stream.

Some devices (and/or processes) in the system will only generate or operate on
a subset of the defined packet formats.  To this end, the protocol is designed
to allow parsers to skip over packets that are unknown or irrelevant to the
task at hand.  For example, the SMS will use the same format for stable storage
as it uses for network traffic. The SMS will place synchronization packets
(section~\ref{section:protocol_sync}) onto stable storage which may be visible
to live-event clients due to the mechanism used for transmission.  Those
clients would ignore the synchronization packets as irrelevant to their
funciton.

\detail{Geometry Information}{%
It is desirable to communicate the geometry of the instrument alongside
the event data generated. This allows Mantid and other tools to stop using
their own geometry files and avoids a potential source of inconsistency.
In order to do this, we will need to define one or more packet type(s) for
this data, as an alternative to the existing post-mortem XML file-based
transmission of this information. There are locations for storing the
geometry information in the NeXus files, but these are specified in
absolute positions and Mantid needs relative position information, so
there will be some standards work here.}

\subsection{Common Protocol Information}

As the bulk of computing power at the SNS is based on the Intel architecture,
all multi-byte values in the protocol will be represented in little-endian
format to avoid the overhead of byte-swapping every value. Any big-endian
hardware added in the future will need to accommodate the existing format.

All packets must be a multiple of 4 bytes in length to keep the integer
fields naturally aligned.

The common field types are given in Table~\ref{table:protocol_field_types}.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{l | l}
	Type & Definition \\
	\hline
	{\bf byte} & 8 bit value, unspecified sign \\
	{\bf word} & 32 bit value, unspecified sign \\
	{\bf u16} & 16 bit unsigned integer \\
	{\bf s32} & 32 bit signed integer \\
	{\bf u32} & 32 bit unsigned integer \\
	{\bf u64} & 64 bit unsigned integer \\
	{\bf u128} & 128 bit unsigned integer \\
	{\bf double} & IEEE 754 64-bit floating point number \\
	{\bf string} & UTF-8 encoded character data of specified length \\
    \end{tabular}
  \end{center}
  \caption {Field Type Definitions}
  \label{table:protocol_field_types}
\end{table}

\subsection{Common Packet Header}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \begin{leftwordgroup}{Header}
      \wordbox{1}{Payload Length (u32)} \\
      \wordbox{1}{Format (u32)} \\
      \wordbox{1}{Timestamp (seconds, u32)} \\
      \wordbox{1}{Timestamp (nanoseconds, u32)}
    \end{leftwordgroup} \\
    \begin{leftwordgroup}{Payload}
      \wordbox[lrt]{1}{Data} \\
      \skippedwords \\
      \wordbox[lrb]{1}{}
    \end{leftwordgroup}
  \end{bytefield}
  \caption{ADARA Network Packet Structure}
  \label{fig:protocol_packet}
\end{figure}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,8,7,0} \\
    \bitbox{24}{Format ID} &
    \bitbox{8}{Version}
  \end{bytefield}
  \caption{Packet Format Fields}
  \label{fig:protocol_format_fields}
\end{figure}


To ensure that parsers are able to find the next packet in the stream, each
packet includes a 16 byte header. This header consists of a payload length,
packet type, and timestamp information as seen in
Figure~\ref{fig:protocol_packet}.

Each packet has an optional payload. If the packet carries information beyond
the header itself, then the {\bf Payload Length} field will be non-zero and a
mulitple of 4. It is expected that stream parsers will first read the full 16
bytes of the header. The {\bf Payload Length} field will then give the length
remaining to be read from the stream for the packet.

Following the {\bf Payload Length}, there is a {\bf Format} field. This 4 byte
field allows for over 4 billion packet types when used sequentially, though the
field is broken into subfields. While the {\bf Format} field may be used
as a plain integer by the parser, the subfields allow some developer
convenience by keeping different versions of packets with the same semantics
numerically close together. The field breakout is shown in
Figure~\ref{fig:protocol_format_fields}. When refering to a format in
this document, the value will be specified by the Format ID, followed by
a period and the version number.

Once a {\bf Format} value has been deployed to production,
the packet format associated with that number shall become immutable. Should
changes be needed to the physical layout or semantic content of that packet, a
new packet type must be allocated and defined in this document. This will
ensure that tools are able to recognize when they are operating on data that
they do not understand.

Table~\ref{table:protocol_format_table} contains the currently assigned
packet types. The Format ID range 0 to 0x3fff is reserved for use by
the detector system, 0x4000 and above are currenly reserved to the ADARA
system.

\begin{table}
  \begin{center}
    \begin{tabular}{l | l | l}
	Type.Version & Name & Section \\
	\hline
	0x0000.0 & Raw Event Data &
		\ref{section:protocol_raw_event_packet} \\
	0x0001.0 & Real Time Data Link Info &
		\ref{section:protocol_rtdl_packet} \\
	0x4000.0 & Banked Event Data &
		\ref{section:protocol_banked_events} \\
	0x4001.0 & Beam Monitor Event Data &
		\ref{section:protocol_beam_monitor_events} \\
	0x4002.0 & Pixel Mapping Table &
		\ref{section:protocol_pixel_mapping_table} \\
	0x4003.0 & Run Status &
		\ref{section:protocol_run_status} \\
	0x4004.0 & Run Informtaion &
		\ref{section:protocol_run_information} \\
	0x4005.0 & Translation Complete &
		\ref{section:protocol_translation_complete} \\
	0x4006.0 & Client Hello &
		\ref{section:protocol_client_hello} \\
	0x4007.0 & Statistics Reset &
		\ref{section:protocol_stats_reset} \\
	0x4008.0 & Synchronization (File) &
		\ref{section:protocol_sync} \\
	0x4009.0 & Heartbeat &
		\ref{section:protocol_heartbeat} \\
	0x8000.0 & Device Descriptor &
		\ref{section:protocol_device_descriptor} \\
	0x8001.0 & Variable Value (u32) &
		\ref{section:protocol_variable_value_u32} \\
	0x8002.0 & Variable Value (double) &
		\ref{section:protocol_variable_value_double} \\
	0x8003.0 & Variable Value (string) &
		\ref{section:protocol_variable_value_string} \\
    \end{tabular}
  \end{center}
  \caption {Assigned Packet Formats}
  \label{table:protocol_format_table}
\end{table}

Following the {\bf Format} field, there is a 64-bit {\bf Timestamp}, broken
into two 32-bit fields. The timestamp gives the seconds (first field) and
nanoseconds since the EPICS epoch, defined as midnight, January 1, 1990.  All
timestamp values are given in UTC time.  This field shall be filled with the
pulse ID given by the accelerator system for all data packets carrying neutron
or ``fast'' environment data. As this field will be used to correlate data from
different sources with the neutron event data, all other packets shall use as
accurate a clock as is available on the local system. This clock should be
closely synchronized to same reference source as the accelerator system.

\newpage
\subsection{Detector System generated packets}

The following packets are generated by the detector system. As such, they will
use the Pulse ID from the Real Time Data Link (RTDL) system to identify each
event an accelerator pulse. Note that while the time-of-flight information for
each pixel has been corrected for frame skew~\footnote{Frame skew indicates a
time-of-flight larger than the intra-pulse time.  For more information on frame
skew, please see the document ``Neutron Event Timing''.}, the ``fast'' metadata
information has no correction applied.

\detail{Detector Control Packets}{%
We have currently defined the packets required to communicate event data from
the detectors to the rest of the aquisition system, but have not defined the
packets that will be used to initialize and control the detector system, as
they are not believed to block progress and the software developers do not
currently have the knowledge required to define them. The detector system
developers will fill them in at a later date.}

\newpage
\subsubsection{Detector System: Raw event data}
\label{section:protocol_raw_event_packet}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{Length $8 N + 16$} \\
    \wordbox{1}{0x00000000} \\
    \wordbox{1}{Pulse ID (seconds)} \\
    \wordbox{1}{Pulse ID (nanoseconds)} \\

    \bitheader{31,30,29,24,23,22,21,16,15,10,9,0} \\
    \wordbox{1}{Source ID} \\
    \bitbox{1}{\tiny E \\ O \\ P} &
    \bitbox{15}{Pkt Seq} &
    \bitbox{16}{DSP Seq} \\
    \bitbox{8}{Flavor} &
    \bitbox{24}{Pulse Charge (units of 10 nC)} \\
    \bitbox{1}{\tiny B \\ V \\ T } &
    \bitbox{1}{\tiny B \\ C \\ Y } &
    \bitbox{8}{TSTAT} &
    \bitbox{12}{Veto} &
    \bitbox{10}{Cycle} \\
    \wordbox{1}{Intra-pulse time (units of 100 ns)} \\
    \bitbox{1}{\tiny R \\ A \\ W} &
    \bitbox{31}{TOF Offset (units of 100 ns)} \\
    \wordbox{2}{Event 1} \\
    \wordbox{2}{$\cdots$} \\
    \wordbox{2}{Event $N$} \\
  \end{bytefield}
  \caption{Packet Type 0x0.0: Raw Event Data}
  \label{fig:protocol_packet_raw_event}
\end{figure}

The ``Raw Event Data'' packet described in
Figure~\ref{fig:protocol_packet_raw_event} is generated by the neutron
detector system and is used to convey time-of-flight and physical location
information for detected neutrons. It contains the following fields:
\begin{itemize}
\item{\bf Length} is defined by the number of events in the packet. There
are 16 bytes of pulse and packet information followed by 8 bytes of data
for each event.
\item{\bf Pulse ID} identifies the most recent pulse generated by the
accelerator. This ID may not be orginating pulse for the events in this packet
due to frame skew.
\item{\bf Source ID} identifies the upstream detector hardware that
processed these events.
\item{\bf EOP} (End of Packet) indicates that this is the last event packet
to be sent for the current pulse.
\item{\bf Pkt Seq} is a monotonically increasing event packet number for the
current pulse.
\item{\bf DSP Seq} is a monotonically increasing packet number for this source.
This field is incremented for every packet sent by the source and does not
reset for each pulse.
\item{\bf Flavor} contains the accelerator pulse flavor as defined by
the SNS Timing Master Functional Description.
\item{\bf Pulse Charge} indicates the charge of the most recent pulse. This
number must be multiplied by 10 to get the charge in nanoColumbs. This
value may not be the charge for the events in this packet due to frame skew.
\item{\bf BVT} indicates that a bad veto frame (frame 24) was received over the
Real Time Data Link.
\item{\bf BCY} indicates that a bad cycle frame (frame 5) was received over the Real Time Data Link.
\item{\bf TSTAT} contains the timing status as defined in the SNS Timing Master
Functional Description.
\item{\bf Veto} contains the veto information for this frame as defined
in the SNS Timing Master Functional Description. A non-zero value indicates
a veto.
\item{\bf Cycle} identifies the Accelerator cycle number for the most
recent pulse. This may not be the originating pulse for the events in this
packet due to frame skew.
\item{\bf Intra-pulse time} contains the intra-pulse time in units of 100 ns.
The intra-pulse time is the time between the previous $T_{SYNC}$ and the
current $T_{SYNC}$.
\item{\bf RAW} indicates that the time of flight values in this packet are
not corrected for frame skew.
\item{\bf TOF Offset} is the total offest that was (or should be, if {\bf RAW}
is set) added to these events to correct for frame skew.
\item{\bf Event(s)} give the time of flight and location of the detected
neutron(s), in a format given by Figure~\ref{fig:protocol_packet_event_data}.
Pixel ID is the physical pixel associated with this neutron event.
\end{itemize}

When run over a network transport that does not provide reliable service, such
as UDP, the protocol uses the {\bf Pulse ID}, {\bf Source ID}, {\bf Pkt Seq},
and {\bf EOP} field to detect the loss of packets describing the events for a
pulse. Each packet source ({\bf Source ID}) sequentially numbers the raw event
packets it sends for a pulse, starting at zero. The source must set {\bf EOP}
on the last raw event packet it sends for a pulse before sending any packet
with a new {\bf Pulse ID}. The receiver will then be able to detect lost
packets by the missing sequence numbers. Should the {\bf EOP} packet be lost,
this will be detected when the sources sends the first packet with a new {\bf
Pulse ID}. This loss-detection protocol assumes that the transport does not
reorder packets; additional measures must be taken in the receiver
to deal with reordering should it be necessary.


\begin{figure}
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{Time-of-flight (units of 100ns, u32)} \\
    \wordbox{1}{Pixel ID (u32)}
  \end{bytefield}
  \caption{Event Data Format}
  \label{fig:protocol_packet_event_data}
\end{figure}

\newpage
{\large \bf Pixel ID Semantics for Raw Event Packets}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,30,28,27,0} \\
    \bitbox{1}{\tiny E \\ R \\ R} &
    \bitbox{3}{Src} &
    \bitbox{28}{Raw ID}
  \end{bytefield}
  \caption{Raw Pixel ID Fields}
  \label{fig:protocol_raw_pixel_id}
\end{figure}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,30,28,27,5,4,3,1,0} \\
    \bitbox{1}{\tiny E \\ R \\ R} &
    \bitbox{3}{Src} &
    \bitbox{23}{Unused} &
    \bitbox{1}{\tiny T \\ R \\ G} &
    \bitbox{3}{Num} &
    \bitbox{1}{\tiny R \\ I \\ S} &
  \end{bytefield}
  \caption{Chopper ID Fields}
  \label{fig:protocol_chopper_id}
\end{figure}

\begin{table}[h]
  \begin{center}
    \begin{tabular}{c | l}
	Value & Source \\
	\hline
	0 & Detector \\
	1,2,3 & Available for future use \\
	4 & Beam Monitor \\
	5 & Special 1 \\
	6 & Special 2 \\
	7 & Chopper \\
    \end{tabular}
  \end{center}
  \caption {Event Source Definitions}
  \label{table:protocol_event_source_id}
\end{table}

The physical pixel ID carried in each event carries semantics according to
Figure~\ref{fig:protocol_raw_pixel_id}:
\begin{itemize}
\item {\bf ERR} indicates an error occurred when set.
\item {\bf Src} indicates the source of this event as defined in
Table~\ref{table:protocol_event_source_id}.
\item {\bf Raw ID} indicates pixel state and ID, chopper variable, beam
monitor ID, or metadata ID, depending on {\bf Src}.
\end{itemize}

When {\bf Src} is 0, indicating that this event was generated by a detector,
{\bf Raw ID} contains both physical pixel ID and state information. The
state information is recovered by dividing the ID value by one more
than the maximum pixel ID for the beamline. The integer portion of the
result is the state number, and the actual physical pixel ID is the
remainder.

When {\bf Src} is 4, {\bf Raw ID} indicates which beam monitor generated
this event.

{\bf Src} values 5 and 6 indicate that this event carries fast metadata
information. The remaining portion of the pixel ID and TOF field currently
remain undefined.

When {\bf Src} is 7, {\bf Raw ID} indicates the chopper and specific variable
contained in the event, according to Figure~\ref{fig:protocol_chopper_id}.
{\bf Num} indicates the chopper number less one (0 is chopper 1, etc.)
{\bf TRG} set indicates that this is a Trigger time; clear indicates that
this is a Top Dead Center time. {\bf RIS} set indicates that this is the
rising edge; clear indicates falling edge.

\detail{Polarization State Information}{%
As part of the ADARA system upgrade, we expect to use fast metadata events to
memoralize transition between polarization states. These will be transformed
into variable update packets once out of the detector system. While this
is the plan, there may be other reasons to keep the current mapping scheme,
which will be determined during implementation.}
\detail{Beam Monitor Mapping}{%
It is assumed, but not yet assured, that the pixel ID for a beam monitor
is in fact the actual beam monitor number. If this is determined not to be
the case, a mapping table will needed.}
\detail{Fast Metadata}{%
There are many formats for the fast metadata pixel IDs, and the pixel ID
also carries the actual value of the metadata variable. The TOF field's
current use is still somewhat unclear to the author, but will contain a
indication of the frame correction applied. During implementaion, effort
will need to be spent documenting and implementing transforms for the
fast metadata pixel ID formats.}

\newpage
\subsubsection{Detector System: Real Time Data Link Information}
\label{section:protocol_rtdl_packet}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{128} \\
    \wordbox{1}{0x00000100} \\
    \wordbox{1}{Pulse ID (seconds)} \\
    \wordbox{1}{Pulse ID (nanoseconds)} \\

    \bitheader{31,30,29,24,23,22,21,10,9,0} \\
    \bitbox{8}{Flavor} &
    \bitbox{24}{Pulse Charge (units of 10 nC)} \\
    \bitbox{1}{\tiny B \\ V \\ T } &
    \bitbox{1}{\tiny B \\ C \\ Y } &
    \bitbox{8}{TSTAT} &
    \bitbox{12}{Veto} &
    \bitbox{10}{Cycle} \\
    \wordbox{1}{Intra-pulse time (units of 100 ns)} \\
    \bitbox{1}{\tiny R \\ A \\ W} &
    \bitbox{31}{TOF Offset (units of 100 ns)} \\
    \wordbox{1}{Ring Period (picoseconds, u32)} \\
    \bitbox{8}{FNA 0} &
    \bitbox{24}{Frame Data 0} \\
    \wordbox{1}{$\cdots$} \\
    \bitbox{8}{FNA 26} &
    \bitbox{24}{Frame Data 26} \\
  \end{bytefield}
  \caption{Packet Type 0x1.0: Real Time Data Link Information}
  \label{fig:protocol_packet_rtdl}
\end{figure}

The ``Real Time Data Link'' packet described in
Figure~\ref{fig:protocol_packet_rtdl} is generated by the neutron
detector system and is used to information about the a given pulse
from the accelerator.  It contains the following fields:
\begin{itemize}
\item{\bf Length} is always 128 bytes.
\item{\bf Pulse ID} identifies the most recent pulse generated by the
accelerator.
\item{\bf Flavor} contains the accelerator pulse flavor as defined by
the SNS Timing Master Functional Description.
\item{\bf Pulse Charge} indicates the charge of the previous pulse. This
number must be multiplied by 10 to get the charge in nanoColumbs.
\item{\bf BVT} indicates that a bad veto frame (frame 24) was received over the
Real Time Data Link.
\item{\bf BCY} indicates that a bad cycle frame (frame 5) was received over the Real Time Data Link.
\item{\bf TSTAT} contains the timing status as defined in the SNS Timing Master
Functional Description.
\item{\bf Veto} contains the veto information for the previous frame as defined
in the SNS Timing Master Functional Description. A non-zero value indicates
a veto.
\item{\bf Cycle} identifies the Accelerator cycle number for the most
recent pulse.
\item{\bf Intra-pulse time} contains the intra-pulse time in units of 100 ns.
The intra-pulse time is the time between the previous $T_{SYNC}$ and the
current $T_{SYNC}$.
\item{\bf RAW} indicates that the time of flight values in this packet are
not corrected for frame skew.
\item{\bf TOF Offset} is the total offest that was (or should be, if {\bf RAW}
is set) added to these events to correct for frame skew.
\item{\bf Ring Period} indicates the number of picoseconds the proton beam
takes to travel the circumference of the accumulator ring. This is used
to calculate beam energy.
\item{\bf FNA $X$} indicates the source RTDL Frame number for
{\bf Frame Data $X$}. A zero in this field indicates no data in
the corresponding frame data field.
\item{\bf Frame Data $X$} contains the data for the RTDL Frame given by
{\bf FNA $X$}.
\end{itemize}


\newpage
\subsection{Stream Management Service generated events}

The following packet types are generally generated by, or used when
communicating with the Stream Management Service (SMS). These packets generally
use the current time-of-day clock for their {\bf Timestamp} fields, though the
Banked Event Data and Beam Monitor Event Data packets will use the Pulse ID
from the associated Raw Event Data.  While all of the defined packet types may
be seen over a network transport, several have meaning only when seen on
permament storage and should be ignored when seen on the network.

{\large \bf Communication with the Translation Service}

The SMS will connect to the Translation Service and send the data recorded
for an experimental run. Prior to sending the event and environmental data
for the run, the SMS will send the following prologue:
\begin{itemize}
\item Pixel Mapping Table (Section~\ref{section:protocol_pixel_mapping_table})
\item Run Status (Section~\ref{section:protocol_run_status})
\item Run Information (Section~\ref{section:protocol_run_information})
\item one or more Device Descriptors and Variable Value packets
(Section~\ref{section:protocol_meta_data_packets})
\end{itemize}
While Variable Value packets will not proceed the Device Descriptor that
defines their meaning, there is no guaranteed ordering among prologue
packets other than that they will be sent prior to event data
(Sections~\ref{section:protocol_banked_events} and
\ref{section:protocol_beam_monitor_events}).

The Translation Service
may also receive Synchronization Packets (Section~\ref{section:protocol_sync})
and Heartbeat Packets (Section~\ref{section:protocol_heartbeat}), both of
which it should discard. These may appear prior to the prologue.

The Translation Service may recieve a Statistics Reset Packet
(Section~\ref{section:protocol_stats_reset}) at any time. It is recommended
that these packets be logged at minimum, though the Translation Service
is free to respond to these packets as it deems appropriate.

At the end of an experiment recording, the SMS will send a Run Status Packet
indicating the end of a run and that all data has been sent. The Translation
Service will finalize its work, and respond with a Translation Complete Packet
(Section~\ref{section:protocol_translation_complete}) indicating sucess or the
error encountered.  The Translation Service need not wait for receipt of the
Run Status Packet from the SMS prior to indicating an error; early notification
is preferred.  Once the Translation Complete Packet is received from the
Translation Service, the SMS will terminate its connection.


{\large \bf Communication with the Live Event Clients}

Upon connection to the SMS, the client shall send a Client Hello
(Section~\ref{section:protocol_client_hello}) to initiate communication
and provide client requests to the SMS. Once the SMS has received the Client
Hello, it will send the following prologue:
\begin{itemize}
\item Pixel Mapping Table (Section~\ref{section:protocol_pixel_mapping_table})
\item Run Status (Section~\ref{section:protocol_run_status})
\item one or more Device Descriptors and Variable Value packets
(Section~\ref{section:protocol_meta_data_packets})
\end{itemize}
While Variable Value packets will not proceed the Device Descriptor that
defines their meaning, there is no guaranteed ordering among prologue
packets other than that they will be sent prior to event data
(Sections~\ref{section:protocol_banked_events} and
\ref{section:protocol_beam_monitor_events}).

If the client's requested starting timestamp is inside an experimental
run, it will also receive a Run Information Packet
(Section~\ref{section:protocol_run_information}).
The client may also receive Synchronization Packets
(Section~\ref{section:protocol_sync})
and Heartbeat Packets (Section~\ref{section:protocol_heartbeat}), both of
which it should discard. These may appear prior to the prologue.

At the end of an experiment recording, the client will receive a Run Status
Packet with its status set to indicate the end of a run.  The client may take
such action as it deems appropriate, but should not respond to the SMS. If the
client chooses to remain connected to the SMS, it will continue to receive
event data. If a new experiment recording is started, the client will receive
the prologue as defined for communication with the Translation Service, and all
subsequent data will be associated with that experimental recording until it
receives another Run Status Packet indicating the completion of the that run.

A live event client  may recieve a Statistics Reset Packet
(Section~\ref{section:protocol_stats_reset}) at any time. It is recommended
that the client reset any analysis it may be performing on the live
stream and discard previous data upon reception.




\newpage
\subsubsection{Banked Event Data}
\label{section:protocol_banked_events}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{Length 16 + Section Lengths} \\
    \wordbox{1}{0x00400000} \\
    \wordbox{1}{Pulse ID (seconds)} \\
    \wordbox{1}{Pulse ID (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \wordbox{1}{Pulse Charge (units of 10 nC, u32)} \\
    \wordbox{1}{Pulse Energy (eV, u32)} \\
    \wordbox{1}{Ring Period (picoseconds, u32)} \\
    \wordbox{1}{Accelerator Cycle} \\
    \wordbox{1}{Flags} \\
    \wordbox{2}{Banked Event Section 1} \\
    \wordbox{1}{$\cdots$} \\
    \wordbox{2}{Banked Event Section $N$} \\
  \end{bytefield}
  \caption{Packet Type 0x4000.0: Banked Event Data}
  \label{fig:protocol_packet_banked_event_data}
\end{figure}

\begin{table}[h]
  \begin{center}
    \begin{tabular}{c | l}
	Bit & Description \\
	\hline
	0 & Pulse had errors \\
	1 & Partial event data \\
	2 & Pulse was veto'd \\
    \end{tabular}
  \end{center}
  \caption {Banked Event Flag Definitions}
  \label{table:protocol_packet_banked_event_flags}
\end{table}

The ``Banked Event Data'' packet described in
Figure~\ref{fig:protocol_packet_banked_event_data} is generated by the
Stream Management Service and is used to convey time-of-flight and logical
location information for detected neutrons. It contains the following fields:
\begin{itemize}
\item{\bf Length} is defined by the number of events in the packet. There
are 16 bytes of pulse and packet information followed by 8 bytes of data
for each event.
\item{\bf Pulse ID} identifies the most recent pulse generated by the
accelerator. This ID may not be orginating pulse for the events in this packet
due to frame skew.
\item{\bf Cycle} identifies the Accelerator cycle number for the most
resent pulse. If this value is unknown (due to an error such as a lost RTDL
packet), this field will be set to all ones. This may not be the originating
pulse for the events in this packet due to frame skew.
\item{\bf Pulse Charge} indicates the charge of the most recent pulse in units
of 10 nanoColumbs. This value may not be the charge for the events in this
packet due to frame skew.
\item{\bf Pulse Energy} indicates the voltage of the most recent pulse,
given in electron volts. This value is calculated from the {\bf Ring Period}
field and depends on the accumulator ring geometry. This value may not be the
voltage for the events in this packet due to frame skew.
\item{\bf Ring Period} indicates the number of picoseconds the proton beam
takes to travel the circumference of the accumulator ring. This is used
to calculate beam energy.
\item{\bf Flags} is a bit-field indicating pulse properties, such as errors or
missing event data as defined in
Table~\ref{table:protocol_packet_banked_event_flags}.
\item{\bf Banked Event Section(s)} contains the Bank ID and events for each
bank. There will be one section for each bank with events for this pulse. The
format of the section is given in
Figure~\ref{fig:protocol_banked_event_section}.
\end{itemize}


While the ``Raw Event Data'' packet described in
section~\ref{section:protocol_raw_event_packet} is defined for primary use over
an unreliable transports and relatively small transmission sizes, the ``Banked
Event Data'' packet is oriented for consumption by analysis systems
communicating over reliable transports, with the ability to send larger
messages. Thus, the Banked packet does not require mechanisms to detect lost
events for a pulse.

The physical layout of pixel IDs in the detector systems for an instrument
do not always match up with the layout that the scientists wish to use for
analysis. The Stream Managment Service will transform the pixel IDs from
physical in the Raw packets to logical in the Banked packets using a
one-to-one mapping (communicated in the stream via ``Pixel Mapping Table''
packets defined in Section~\ref{section:protocol_pixel_mapping_table}.
These logical pixels are then communicated in the ``Banked Event Data''
packets. As part of this mapping process, the detector bank is determined,
and the events are sorted into buckets accordingly. It is expected that
a single ``Banked Event Data'' packet will be sent by the SMS for each pulse.

\begin{figure}
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{Bank ID} \\
    \wordbox{1}{Event Count} \\
    \wordbox{2}{Banked Event 1} \\
    \wordbox{1}{$\cdots$} \\
    \wordbox{2}{Banked Event $N$} \\
  \end{bytefield}
  \caption{Banked Event Section}
  \label{fig:protocol_banked_event_section}
\end{figure}

Each ``Banked Event Data'' packet contains one ``Banked Event Section'' for
each detector bank with events in this pulse. As defined in
Figure~\ref{fig:protocol_banked_event_section}, each section contains
the following fields:
\begin{itemize}
\item{\bf Bank ID} indicates which detector bank is the source of the
events in this section.
\item{\bf Event Count} indicates the number of events in this section. This
may be zero if there were no events for this pulse from this detector bank.
\item{\bf Event(s)} give the time of flight and location of the detected
neutron(s), in a format given by Figure~\ref{fig:protocol_packet_event_data}.
Pixel ID is the logical pixel ID, translated from the raw ID given by the
detector.
\end{itemize}

\detail{Define Banked Event Data Flags field}{%
We need a flag to indicate that we may be missing data from this frame.
We may need a veto flag, and perhaps some of this space should be used
to indicate polarization state, though I believe that should be handled
via a variable update packet.
We have space here, are there other flags that make sense in this context?}
\detail{Use additional information from RTDL}{%
We now have the intra-pulse time and other fields we will want to record
in this packet. However, can different pixels in the same detector bank
get different TOF offsets? If so, that will cause a problem when mixing
those. The intra-pulse time will be OK, but what about $T_{DELAY}$? If
$T_{DELAY}$ is in the reported TOF, but not in the intra-pulse time, then
we should be OK, as we the intra-pulse time is the same for all events.
Resolve this for implementation.}

\newpage
\subsubsection{Beam Monitor Event Data}
\label{section:protocol_beam_monitor_events}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{Length 16 + Section Lengths} \\
    \wordbox{1}{0x00400100} \\
    \wordbox{1}{Pulse ID (seconds)} \\
    \wordbox{1}{Pulse ID (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \wordbox{1}{Pulse Charge (units of 10 nC, u32)} \\
    \wordbox{1}{Pulse Energy (eV, u32)} \\
    \wordbox{1}{Ring Period (picoseconds, u32)} \\
    \wordbox{1}{Accelerator Cycle} \\
    \wordbox{1}{Flags} \\
    \wordbox{2}{Beam Monitor Section 1} \\
    \wordbox{1}{$\cdots$} \\
    \wordbox{2}{Beam Monitor Section $N$} \\
  \end{bytefield}
  \caption{Packet Type 0x4001.0: Beam Monitor Event Data}
  \label{fig:protocol_packet_monitor_event_data}
\end{figure}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,24,23,0} \\
    \bitbox{8}{Monitor ID} &
    \bitbox{24}{Event Count} \\
    \wordbox{1}{Time-of-flight 1 (units of 100 ns, u32)} \\
    \wordbox{1}{$\cdots$} \\
    \wordbox{1}{Time-of-flight $N$} \\
  \end{bytefield}
  \caption{Beam Monitor Section}
  \label{fig:protocol_monitor_event_section}
\end{figure}

The ``Beam Monitor Event Data'' packet described in
Figure~\ref{fig:protocol_packet_monitor_event_data} is generated by the
Stream Management Service and is used to convey neutron time-of-flight
information originating from beam monitors. It contains the following fields:
\begin{itemize}
\item{\bf Length} is defined by the number of events in the packet. There
are 16 bytes of pulse and packet information followed by 4 bytes of data
for each event.
\item{\bf Pulse ID} identifies the most recent pulse generated by the
accelerator. This ID may not be originating pulse for the events in this packet
due to frame skew.
\item{\bf Cycle} identifies the Accelerator cycle number for the most
recent pulse. If this value is unknown (due to an error such as a lost RTDL
packet), this field will be set to all ones. This may not be the originating
pulse for the events in this packet due to frame skew.
\item{\bf Pulse Charge} indicates the charge of the most recent pulse in units
of 10 nanoColumbs. This value may not be the charge for the events in this
packet due to frame skew.
\item{\bf Pulse Energy} indicates the voltage of the most recent pulse,
given in electron volts. This value is calculated from the {\bf Ring Period}
field and depends on the accumulator ring geometry. This value may not be the
voltage for the events in this packet due to frame skew.
\item{\bf Ring Period} indicates the number of picoseconds the proton beam
takes to travel the circumference of the accumulator ring. This is used
to calculate beam energy.
\item{\bf Flags} is a bit-field indicating pulse properties, such as errors or
missing event data as defined in
Table~\ref{table:protocol_packet_banked_event_flags}.
\item{\bf Beam Monitor Section(s)} contain the Beam Monitor ID and events
for each Beam Monitor. There will be one section for each known beam monitor
on this beam line that has events for this pulse. The format of the section
is given in Figure~\ref{fig:protocol_monitor_event_section}.
\end{itemize}

While the ``Raw Event Data'' packet described in
section~\ref{section:protocol_raw_event_packet} is defined for primary use over
an unreliable transports and relatively small transmission sizes, the ``Beam
Monitor Event Data'' packet is oriented for consumption by analysis systems
communicating over reliable transports, with the ability to send larger
messages. Thus, the Beam Monitor packet does not require mechanisms to detect
lost events for a pulse.

Each ``Beam Monitor Event Data'' packet contains one ``Beam Monitor Section''
for each Beam Monitor that has events in this pulse. As defined in
Figure~\ref{fig:protocol_monitor_event_section}, each section contains the
following fields:
\begin{itemize}
\item{\bf Monitor ID} indicates which beam monitor is the source of the events
in this section.
\item{\bf Event Count} indicates the number of events in this section. This
may be zero if there were no events for this pulse from this beam monitor.
\item{\bf Time-of-flight} give the time-of-flight in units of 100 nanoseconds
of the detected neutrons.
\end{itemize}

Beam monitors do not record pixel IDs to record the location of detected
neutrons, so there is no need to spend the space storing that information.
This is especially important as the monitors may often be exposed to ``white
beam'' conditions, generating non-trivial amounts of data in spite of their
low detection efficiency.

It is expected that a single ``Beam Monitor Event Data'' packet will be sent
by the SMS for each pulse.

\detail{Define Beam Monitor Event Data Flags field}{%
We need a flag to indicate that we may be missing data from this frame.
We may need a veto flag, and perhaps some of this space should be used
to indicate polarization state, though I believe that should be handled
via a variable update packet.
We have space here, are there flags that make sense in this context?}

\newpage
\subsubsection{Pixel Mapping Table Packet}
\label{section:protocol_pixel_mapping_table}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{$N$} \\
    \wordbox{1}{0x00400200} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,0} \\
    \wordbox{2}{Mapping Data ($N$ bytes)} \\
  \end{bytefield}
  \caption{Packet Type 0x4002.0: Pixel Map Packet}
  \label{fig:protocol_packet_pixel_map}
\end{figure}

As part of its processing of the neutron event data, the SMS will map the
physical pixel ID to a logical pixel ID. This mapping must be one-to-one and
reversible. The ``Pixel Mapping Table'' packet defined in
Figure~\ref{fig:protocol_packet_pixel_map} will preserve the table for
consumers of the event stream. This packet must be present near the beginning
of a new experiment recording, but may occur multiple times in the transmitted
stream. Each additional copy transmitted must be identical to the first copy
for the same experiment recording.

\begin{figure}[ht]
  \centering
  \begin{bytefield}{48}
    \bitheader{47,16,15,0} \\
    \bitbox{32}{Base Physical ID (u32)} &
    \bitbox{16}{Count (u16)} \\
    \wordbox{1}{Map Entry 0} \\
    \wordbox[lrt]{1}{Map Entry 1} \\
    \wordbox[lr]{1}{$\cdots$} \\
    \wordbox[lrb]{1}{Map Entry $N$} \\
  \end{bytefield}
  \caption{Pixel Map Section}
  \label{fig:protocol_packet_pixel_map_section}
\end{figure}

\begin{figure}[ht]
  \centering
  \begin{bytefield}{48}
    \bitheader{47,16,15,0} \\
    \bitbox{32}{Logical ID (u32)} &
    \bitbox{16}{Bank ID (u16)} \\
  \end{bytefield}
  \caption{Pixel Map Entry}
  \label{fig:protocol_packet_pixel_map_entry}
\end{figure}

The physical pixel ID space may not be compactly populated. It often
contains sequential clusters of IDs with large gaps between the clusters.
To reduce the size of the mapping table, it is described using sections as
defined in Figure~\ref{fig:protocol_packet_pixel_map_section}.
{\bf Base Physical ID} defines be offset into the physical ID space for
which this section defines mappings. {\bf Count} gives the number of
mapping entries in this section.

The mapping entries are defined in
Figure~\ref{fig:protocol_packet_pixel_map_entry}.  The physical pixel ID for
the entry is determined by adding its zero-based index into the section to the
{\bf Base Physical ID}. This physical ID is then mapped to the {\bf Logical ID}
specified, and was generated by a detector in bank {\bf Bank ID}.

The final section of the table may have 2 zero bytes appended to
pad the total packet length to a multiple of 4 bytes.


\newpage
\subsubsection{Run Status Packet}
\label{section:protocol_run_status}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{12} \\
    \wordbox{1}{0x00400300} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,24,23,0} \\
    \wordbox{1}{Run Number(u32)} \\
    \wordbox{1}{Run Start (seconds, u32)} \\
    \bitbox{8}{Status} &
    \bitbox{24}{File Number} \\
  \end{bytefield}
  \caption{Packet Type 0x4003.0: Run Status Packet}
  \label{fig:protocol_packet_status}
\end{figure}

\begin{table}[h]
  \begin{center}
    \begin{tabular}{c | l}
	Value & Definition \\
	\hline
	0 & No current run \\
	1 & Beginning new run \\
	2 & Continuing current run (end of file marker) \\
	3 & Continuing current run (new file marker) \\
	4 & Ending current run \\
    \end{tabular}
  \end{center}
  \caption {Run Status Definitions}
  \label{table:protocol_run_status_values}
\end{table}

The ``Run Status'' packet is used to indicate if we are currently in an
experimental run, and if so, identify it. This packet must be sent/written at
the start of each experimental run, and should be written to the start of
each event data file to tag it with the run data it contains.

\begin{itemize}
\item{\bf Length} is 12 bytes.
\item{\bf Timestamp} is the current system time when the packet is created.
\item{\bf Run Number} is the identifier for the current experimental run. This
field is 0 if we are not currently recording an experimental run.
\item{\bf Run Start} contains the seconds portion of the time that we began
the current experiment run. This field is 0 if we are not currently recording
data.
\item{\bf Status} is defined in Table~\ref{table:protocol_run_status_values}.
\item{\bf File Number} indentifies which file this data is stored in. This
field is undefined in we are not currently recording an experimental run.
\end{itemize}

The Stream Management Service breaks experimental recordings into multiple
chunks to allow a gradual reclaimation of storage space once the recording
has been translated to a NeXus file. This process increases the window in
which at least partial data from a experiment can be retrieved from the SMS
in the case that a problem is detected after the guaranteed recovery period.
Rather than relying solely on the filename (which could be subject to human
error) to determine how much of the recording has been preserved, the
``Run Status'' packet provides an authoratative source for the amount
of time and approximate data missing by using the {\bf Run Start} and
{\bf File Number} fields along with the administrator controlled chunk size.

\newpage
\subsubsection{Run Information Packet}
\label{section:protocol_run_information}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{4 + roundup($N$, 4)} \\
    \wordbox{1}{0x00400400} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,0} \\
    \wordbox{1}{XML Length (u32)} \\
    \wordbox{4}{XML Run Information ($N$ byte string data)} \\
  \end{bytefield}
  \caption{Packet Type 0x4004.0: Run Information Packet}
  \label{fig:protocol_packet_runinfo_run}
\end{figure}

\detail{Define Run Information packet}{%
Jim and Marie will define the XML schema for this packet, and we
will reference it from the protocol document. User comments
and other information that is not specifically related will be moved to
metadata value packets.}


\newpage
\subsubsection{Translation Complete Packet}
\label{section:protocol_translation_complete}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{4 + roundup($N$, 4)} \\
    \wordbox{1}{0x00400500} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \bitbox{16}{Status (u16)} &
    \bitbox{16}{Reason Length (u16)} \\
    \wordbox{2}{Reason ($N$ byte string, optional)} \\
  \end{bytefield}
  \caption{Packet Type 0x4005.0: Translation Complete Packet}
  \label{fig:protocol_packet_translation_complete}
\end{figure}

\begin{table}[h]
  \begin{center}
    \begin{tabular}{c | l}
	Value & Definition \\
	\hline
	0 & Success \\
	0x0001 to 0x7fff & Transient Error \\
	0x8000 to 0xffff & Permament Error \\
    \end{tabular}
  \end{center}
  \caption {Status Definitions}
  \label{table:protocol_completion_status_values}
\end{table}

The ``Translation Complete'' Packet allows the Translation Service to signal
the Stream Management Service that is has completed processing the data sent
for this experimental run.

\begin{itemize}
\item{\bf Length} is 8 bytes plus the length of the optional {\bf Reason} field
rounded up to the next multiple of four.
\item{\bf Timestamp} is the current system time when the packet is created.
\item{\bf Status} indicates the result of translation as defined in
Table~\ref{table:protocol_completion_status_values}.
\item{\bf Reason Length} is the length of the comment that follows, not
including padding.
\item{\bf Reason} is an optional UTF-8 encoded reason. This may be used
to communicate a small amount of human readable information to explain
a non-successful status. This field is padded by zeros to the nearest
multiple of 4.
\end{itemize}

If the {\bf Status} contains a value defined to be a Transient Error, the SMS
should attempt to resend this experimental run to the Translation Service.  If
the {\bf Status} contains a value defined to be a Permament Error, the SMS
should not resend this experimental run until a human has inspected and
corrected the root cause. Only if {\bf Status} is successful may the SMS
consider the experimental run safely on disk at the Translation Service and
mark the storage space valid for eventual reclaimation.

The Translation Service may send this packet immediately upon detecting an
error and need not wait for the SMS to signal the end of data for the
experiment.


\newpage
\subsubsection{Client Hello Packet}
\label{section:protocol_client_hello}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{4} \\
    \wordbox{1}{0x00400600} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,0} \\
    \wordbox{1}{Requested Start Timestamp (seconds, u32)} \\
  \end{bytefield}
  \caption{Packet Type 0x4006.0: Client Hello Packet}
  \label{fig:protocol_client_hello}
\end{figure}

After first connecting to the Stream Management Service, a client will send
a ``Client Hello'' packet to indicate from where in the stream it would like
begin receiving event data. No data will be sent to the client prior to
receiving this packet.

\begin{itemize}
\item{\bf Length} is 4 bytes.
\item{\bf Timestamp} is the current system time when the packet is created.
\item{\bf Requested Start Timestamp} suggests to the server where in the live
stream to start sending events to this client, as defined below.
\end{itemize}

The value given in {\bf Requested Start Timestamp} is a suggestion to the
server and may be ignored. The client may receive events preceeding its
requested timestamp, or may only receive live data if historical data is not
available.

The requested value is bounded by run state transitions. If the client
requests data from before the current run started, the server will start
sending events from the beginning of the run. If there is no current run and
the requested timestamp is before the end of the previous run, the server will
begin sending events from immediately after the previous run ended.

The requested timestamp may be rounded to the granularity of the event indexing
used by the Stream Management Service. In this case, the timestamp will be
rounded towards the historical data, and the client may receive events that
preceed the requested point in time. The client is responsible for filtering
the undesired requests from the the stream.

If the {\bf Requested Start Timestamp} is zero, the server will not send
historical data and will begin sending events with the next pulse.
Note, therefore, that to request a replay of all historical data for
a currently active run, the {\bf Requested Start Timestamp} could simply
be set to ``1'' to indicate the oldest possible timestamp greater than 0.

\newpage
\subsubsection{Statistics Reset Packet}
\label{section:protocol_stats_reset}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{0} \\
    \wordbox{1}{0x00400700} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)}
  \end{bytefield}
  \caption{Packet Type 0x4007.0: Statistics Reset Packet}
  \label{fig:protocol_packet_stats_reset}
\end{figure}

During manual operation of a beamline such as performing sample alignment, the
scientist/user may not be recording neutron event data for translation. In
these situations, the user may wish to reset the live histogram generation and
any other temporary or partial analysis result between each phase of the
process. To allow ease of use and manipulation, the live tools should be able
to receive this request via the same data stream as the experiment data. This
allows precise, common boundaries between data sets to be aligned among all of
the live tools, and avoids the need to have the user manipulate multiple tools
to perform the reset.

When non-translation clients receive a ``Statistics Reset Packet'' (defined in
Figure~\ref{fig:protocol_packet_stats_reset}) in the data stream, they should
discard previous neutron data. The Translation Service may note the occurance
of these packets in the appropriate place.

The contents of the packet field are as follows:
\begin{itemize}
\item{\bf Length} is 0 bytes.
\item{\bf Timestamp} is the current system time when the packet is created.
\end{itemize}


\newpage
\subsubsection{Synchronization Packet}
\label{section:protocol_sync}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{28 + roundup($N$, 4)} \\
    \wordbox{1}{0x00400800} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,0} \\
    \wordbox{4}{Signature (16 bytes)} \\
    \wordbox{2}{File Offset (u64)} \\
    \wordbox{1}{Comment Length (u32)} \\
    \wordbox{2}{Comment ($N$ byte string, optional)} \\
  \end{bytefield}
  \caption{Packet Type 0x4008.0: Synchronization Packet}
  \label{fig:protocol_packet_sync}
\end{figure}

While the ADARA protocol is primarily used for network transmission of
event data, it is necessary to store the raw data on storage media during
processing and for archival storage. The ``Synchronization`` packet defined
in Figure~\ref{fig:protocol_packet_sync} allows partial recovery of experiment
data in the event that a media error or other corruption destroys a packet
header. Without a marker in the data stream, it may be impossible to find
the beginning of the next packet, rendering all data after the media defect
unusable. Periodically placing synchronization packets in the stream contains
the damage that the error can cause.

\begin{itemize}
\item{\bf Length} is 28 bytes plus the length of the optional comment field,
rounded up to the nearest multiple of 4 bytes.
\item{\bf Timestamp} is the current system time when the packet is stored
to disk.
\item{\bf Signature} is the byte string 0x53 0x4e 0x53 0x41 0x44 0x41 0x52
0x41 0x4f 0x52 0x4e 0x4c 0x00 0x00 0xf0 0x7f.
\item{\bf File Offset} is the offset from the beginning of the file to the
length field of this packet. This is used to further verify that we have a
valid Synchronization Packet.
\item{\bf Comment Length} is the length of the comment that follows, not
including padding.
\item{\bf Comment} is an optional UTF-8 encoded comment. This may be used
to store a small amount of human readable information that may be displayed
by the Unix `file' command. This field is padded by zeros to the nearest
multiple of 4.
\end{itemize}

Every file containing ADARA data should begin with a Synchronization packet to
allow system utilities to identify the file as such. Additional synchronization
packets should be added to the file as desired to contain any potential damage.
As the Synchonization packet is only useful when used for storage, it should be
ignored when encountered over a network transport.

{\bf Signature} was constructed using the string ``SNSADARA'' and concatenating
a positive, signaling NaN value in IEEE 754 double format. The mantessa bits
marked as ``don't care'' in the standard have the string ``ORNL'' inserted such
that it views properly when displayed in byte order.  As it is expected that
NaN values are unlikely to be part of a valid data stream, this signature
should prove unique when combined with a matching file offset.


\newpage
\subsubsection{Heartbeat Packet}
\label{section:protocol_heartbeat}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{0} \\
    \wordbox{1}{0x00400900} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)}
  \end{bytefield}
  \caption{Packet Type 0x4009.0: Heartbeat Packet}
  \label{fig:protocol_packet_heartbeat}
\end{figure}

In order for clients to distinguish between a crashed server and one with
no event data to transmit (ie, shutter is closed), there must be periodic
traffic over the network transport. The ``Heartbeat'' packet fills this need.

This role may also be performed by the network transport's keepalive settings,
though that does not help communication with the detector system which is
expected to use an unreliable transport.

\newpage
\subsection{Meta Data Packets}
\label{section:protocol_meta_data_packets}

In addition to the neutron events from the detector systems, the SNS must
collect information about the environment of the sample being studied. This
data comes from a diverse and dynamic collection of devices, and requires
a flexible format to communicate their properties. The ADARA network
protocol addresses this requirement by separating information about devices
into two pieces: Device Descriptors and Variable Values.

Device Descriptors contain an XML description of the device. This description
includes information such as the name of the device, which application (and
version thereof) controls it, and a list of variables that the device exports.
Each variable listed in the Device Descriptor will include information about
that variable such as the variable ID, source EPICS process variable name, a
``friendly'' name for the NeXus file and analysis, type (string, integer,
IEEE754 double, etc.), units of measurement, and valid range.

Each Veriable Value Packet contains a Device ID to link the value back
to a specific Device Descriptor, and a Variable ID to identify the variable
being updated. It includes information on the status of the variable
and a seveity of any alarm such that operation outside of design parameters
and data drop outs due to malfunctioning devices may be identified and
the scientist can make an informed decision regarding the suitability
of event data for the affected periods of time.

While allocation of Device IDs is outside the scope of this document, they
will be statically assigned, and coordination of allocations is a must
to avoid duplicate IDs. Device IDs less than 8192 are reserved for
internal ADARA use to report on system status information such as missing
or error packets from the detectors. Other Device IDs will be allocated
to designate certain semi-fixed equipment on the beamlines, such as
choppers.


\newpage
\subsubsection{Variable Status and Severity}

Each variable carries with it an associated alarm status and severity. These
values are derived from the EPICS Record Reference Manual.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{c | l | l}
	Value & EPICS Name & Definition \\
	\hline
	0 & None & No alarm \\
	1 & Read & Unable to read value \\
	2 & Write & Unable to write value \\
	3 & HiHi & Analog value is above the high ``high'' limit \\
	4 & High & Analog value is above the low ``high'' limit \\
	5 & LoLo & Analog value is below the low ``low'' limit \\
	6 & Low & Analog value is below the high ``low'' limit \\
	7 & State & Record is considered to be in bad state \\
	8 & Cos & Digital value changed state \\
	9 & Comm & Device not communicating \\
	10 & Timeout & Device communications timed out \\
	11 & HwLimit & Device indicated hardware limit \\
	12 & Calc & Bad calculation \\
	13 & Scan & Invalid SCAN field \\
	14 & Link & Failed to link to another record \\
	15 & Soft & Record is in invalid state (software error) \\
	16 & BadSub & Subroutine not found \\
	17 & UDF & Alarm status is undefined \\
	18 & Disable & Record was disabled \\
	19 & Simm & Record is in simulation mode \\
	20 & ReadAcess & No Read Access \\
	21 & WriteAccess & No Write Access \\
    \end{tabular}
  \end{center}
  \caption {Variable Status Definitions}
  \label{table:protocol_variable_status_values}
\end{table}

\begin{table}[h]
  \begin{center}
    \begin{tabular}{c | l}
	Value & Definition \\
	\hline
	0 & No Alarm \\
	1 & Minor \\
	2 & Major \\
	3 & Invalid \\
    \end{tabular}
  \end{center}
  \caption {Variable Alarm Severity Definitions}
  \label{table:protocol_variable_severity_values}
\end{table}


\newpage
\subsubsection{Device Descriptor Packet}
\label{section:protocol_device_descriptor}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{8 + roundup($N$, 4)} \\
    \wordbox{1}{0x00800000} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \wordbox{1}{Device ID (u32)} \\
    \wordbox{1}{Descriptor Length (u32)} \\
    \wordbox[lrt]{1}{XML Descriptor ($N$ byte string)} \\
    \skippedwords \\
    \wordbox[lrb]{1}{}
  \end{bytefield}
  \caption{Packet Type 0x8000.0: Device Descriptor Packet}
  \label{fig:protocol_packet_device_desc}
\end{figure}

\begin{itemize}
\item{\bf Length} is 8 bytes plus the length of the {\bf XML Descriptor} field
rounded up to the next multiple of four.
\item{\bf Timestamp} is the current system time when the packet is created.
\item{\bf Device ID} indicates the device this descriptor describes.
\item{\bf Descriptor Length} is the length of the XML descriptor that follows,
not including padding.
\item{\bf XML Descriptor} is a UTF-8 encoded XML document describing the
properties of this variable. This field is padded by zeros to the nearest
multiple of 4.
\end{itemize}

The value in the {\bf Device ID} field is used to identify a device in the
ADARA, and to provide separate variable namespaces for each one.

\detail{Define Device Descriptor XML Schema}{%
Jim and Marie will define the schema for the XML content of this packet.}

\newpage
\subsubsection{Variable Value (u32) Packet}
\label{section:protocol_variable_value_u32}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{12} \\
    \wordbox{1}{0x00800100} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \wordbox{1}{Device ID (u32)} \\
    \wordbox{1}{Variable ID (u32)} \\
    \bitbox{16}{Status} &
    \bitbox{16}{Severity} \\
    \wordbox{1}{Variable Value (u32)}
  \end{bytefield}
  \caption{Packet Type 0x8001.0: Variable Value (u32) Packet}
  \label{fig:protocol_packet_value_u32}
\end{figure}

\begin{itemize}
\item{\bf Length} is 12 bytes.
\item{\bf Timestamp} indicates the time that the variable changed.
\item{\bf Device ID} indicates the device this variable is from.
\item{\bf Variable ID} indicates the variable that changed.
\item{\bf Status} of this variable from
Table~\ref{table:protocol_variable_status_values}
\item{\bf Severity} of this variable from
Table~\ref{table:protocol_variable_severity_values}
\item{\bf Variable Value} is an unsigned 32 bit value for this variable.
\end{itemize}

Note that this packet type is also used to carry enumerated and boolean values.
The specific interpretation of the value is dictated by the Device Descriptor
associated with the Device ID.

\newpage
\subsubsection{Variable Value (double) Packet}
\label{section:protocol_variable_value_double}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{16} \\
    \wordbox{1}{0x00800200} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \wordbox{1}{Device ID (u32)} \\
    \wordbox{1}{Variable ID (u32)} \\
    \bitbox{16}{Status} &
    \bitbox{16}{Severity} \\
    \wordbox{2}{Variable Value (IEEE 754 double)}
  \end{bytefield}
  \caption{Packet Type 0x8002.0: Variable Value (double) Packet}
  \label{fig:protocol_packet_value_double}
\end{figure}

\begin{itemize}
\item{\bf Length} is 16 bytes.
\item{\bf Timestamp} indicates the time that the variable changed.
\item{\bf Device ID} indicates the device this variable is from.
\item{\bf Variable ID} indicates the variable that changed.
\item{\bf Status} of this variable from
Table~\ref{table:protocol_variable_status_values}
\item{\bf Severity} of this variable from
Table~\ref{table:protocol_variable_severity_values}
\item{\bf Variable Value} is an IEEE 754 double floating point value for this
variable.
\end{itemize}

\newpage
\subsubsection{Variable Value (string) Packet}
\label{section:protocol_variable_value_string}

\begin{figure}[h]
  \centering
  \begin{bytefield}{32}
    \bitheader{31,0} \\
    \wordbox{1}{12 + roundup($N$, 4)} \\
    \wordbox{1}{0x00800300} \\
    \wordbox{1}{Timestamp (seconds)} \\
    \wordbox{1}{Timestamp (nanoseconds)} \\

    \bitheader{31,16,15,0} \\
    \wordbox{1}{Device ID (u32)} \\
    \wordbox{1}{Variable ID (u32)} \\
    \bitbox{16}{Status} &
    \bitbox{16}{Severity} \\
    \wordbox{1}{String Length (u32)} \\
    \wordbox{2}{Variable Value ($N$ byte string data)} \\
  \end{bytefield}
  \caption{Packet Type 0x8003.0: Variable Value (string) Packet}
  \label{fig:protocol_packet_value_string}
\end{figure}

\begin{itemize}
\item{\bf Length} is 12 bytes plus the length of the {\bf Variable Value} field
rounded up to the next multiple of four.
\item{\bf Timestamp} indicates the time that the variable changed.
\item{\bf Device ID} indicates the device this variable is from.
\item{\bf Variable ID} indicates the variable that changed.
\item{\bf Status} of this variable from
Table~\ref{table:protocol_variable_status_values}
\item{\bf Severity} of this variable from
Table~\ref{table:protocol_variable_severity_values}
\item{\bf String Length} is the length of the string that follows,
not including padding.
\item{\bf Variable Value} is a UTF-8 encoded string containing the new value
for this variable. This field is padded by zeros to the nearest multiple of 4.
\end{itemize}


